---
title: "Stream Crossing Risk Model"
author: "Chris Hoover"
date: "April 2, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("googledrive_read_write_functions.R")

require(lubridate)
require(tidyverse)
```

# Risk Model from Survey Data  
Available data on stream crossing risk is provided from John Ladd's JMT survey as a rating from 0-5 describing perceived difficulty of all stream crossings during the individual's trip. The stream crossing risk model will use this rating as the outcome, classifying any rating >1 as "risky" and any rating <=1 as not risky (to generate binary outcome). Covariates in the model will include individual hiker characteristics including: weight, height, age, gender and start date 

```{r load_merge_data, message=FALSE, warning=FALSE}
daily_snow_melt <- load_csv_from_googledrive("12qFuFl2Vj3jhJiGjjdviqT6xKw3oprZn")
person_crossings <- load_csv_from_googledrive("12P_z60jhEPqWLpYr23l1w3fFZlMAzSbh")
person_covariates <- load_csv_from_googledrive("1CmPS7bhIHSsy-6dkjfsSuZmYoYXcGOe8")

merged_data <- person_crossings %>% 
  left_join(person_covariates, by = c("person_id" = "UniqueID")) %>% 
  left_join(daily_snow_melt, by = c("crossing_name" = "watershed",
                                    "crossing_date" = "Date")) 
model_data <- merged_data %>% 
  group_by(person_id) %>% 
  summarise(overall_stream_challenge = first(overall_stream_challenge),
            peak_SWE_melt = max(SWE_melt),
            mean_SWE_melt = mean(SWE_melt),
            peak_melt_crossing = ifelse(peak_SWE_melt == 0, "Not Applicable", 
                                        crossing_name[which(SWE_melt == peak_SWE_melt)]),
            peak_SWE = max(SWE),
            mean_SWE = mean(SWE),
            peak_SWE_crossing = ifelse(peak_SWE == 0, "Not Applicable", 
                                       crossing_name[which(SWE == peak_SWE)]),
            gender = first(`Gender (compiled and best evidence)`),
            age = first(`Age at entry TH`),
            height_in = as.numeric(first(`What is your height? feet_y`))*12 + 
                        as.numeric(first(`What is your height? inches_y`)),
            weight = first(`What did you weigh at the start and end of your hike? (Estimates are acceptable.) Starting body weight`),
            MPD = first(`Calc mean MPD`),
            Group_Size = first(`GrpSize AQ19  / Not counting any friends you first met on your hike, what was your group size when at its largest.`),
           # Experience = first(),
            StartDate = as.Date(first(StartDate)),
            start_week = week(StartDate)) %>% 
  mutate(crossing_difficulty_binary = ifelse(overall_stream_challenge <= 1, 0, 1),
         crossing_difficulty_prop = overall_stream_challenge / 5)
```

```{r descriptive}
table(model_data$overall_stream_challenge)

table(model_data$peak_SWE_crossing)
```

Individuals whose peak snow melt across all crossings was 0 (implying there was no snow melt) were labeled as "Not Applicable". Among those who did experience snow melt at crossings, most appear to experience it at Ireland Creek, Hilgard Branch, South Fork Kings, Bear, and Bubbs creeks

```{r more descriptive}
model_data %>% 
  filter(peak_SWE_melt > 0) %>% 
  ggplot(aes(peak_SWE_melt)) + geom_histogram() + theme_classic()

model_data %>% 
  ggplot(aes(start_week)) + geom_histogram() + theme_classic() 

model_data %>% 
  ggplot(aes(height_in)) + geom_histogram() + theme_classic() 

model_data %>% 
  ggplot(aes(age)) + geom_histogram() + theme_classic() 

```


```{r init_bin_mod}
crossing_difficulty_bin_mod <- glm(crossing_difficulty_binary ~ peak_SWE_melt,
                               family = "binomial", data = model_data)
  summary(crossing_difficulty_bin_mod)
```

Very small, but highly significant increase in odds of rating streams as >1 associated with increased now melt experienced. Let's look at a linear model, which assumes that rating 0-5 is continuous (which we know isn't true) but is worth checking out regardless.

```{r init_lin_mod}
model_data %>% 
  ggplot(aes(x = peak_SWE_melt, y = overall_stream_challenge, col = peak_melt_crossing)) +
    geom_point() +
    theme_classic() +
    theme(legend.position = "bottom")

crossing_difficulty_lin_mod <- lm(overall_stream_challenge ~ peak_SWE_melt,
                               data = model_data)
  summary(crossing_difficulty_lin_mod)

```

```{r adjust_bin_mod}
crossing_difficulty_bin_mod_adj <- glm(crossing_difficulty_binary ~ peak_SWE_melt + gender + age + height_in + weight + start_week,
                               family = "binomial", data = model_data)
  summary(crossing_difficulty_bin_mod_adj)
```

```{r adjust_linin_mod}
crossing_difficulty_lin_mod_adj <- lm(overall_stream_challenge ~ 
                                        peak_SWE_melt + gender + age + 
                                        height_in + weight + start_week + 
                                        MPD + Group_Size,
                                      data = model_data)
  summary(crossing_difficulty_lin_mod_adj)
  
save(crossing_difficulty_lin_mod_adj, file = "risk_model_object.Rdata")  
```

```{r adjust_linin_mod_standardize}
crossing_difficulty_lin_mod_adj <- lm(overall_stream_challenge ~ 
                                        scale(peak_SWE_melt) + gender + scale(age) + 
                                        scale(height_in) + scale(weight) + scale(start_week) + 
                                        scale(MPD) + scale(Group_Size),
                                      data = model_data)
  summary(crossing_difficulty_lin_mod_adj)
```

# Temp-melt model  
In order to do any sort of predictions, have to be able to relate input variables to variables that are forecasted (like temperature). Attampe here to model snow melt in watersheds as a function of antecedent SWE and temperature  

```{r temp_melt}
daily_temp <- load_csv_from_googledrive("1GAmIE-5zCUjrLJebs4G-u9mT0qniL6B4")

temp_melt_merge <- daily_temp %>% 
  left_join(daily_snow_melt %>% filter(Date <= ymd("2016-01-01")), by = c("Date" = "Date",
                                                                          "watershed" = "watershed"))

temp_melt_merge %>% ggplot(aes(x = Mean_Temp, y = SWE_melt, col = watershed)) +
  theme_classic() +
  geom_point() +
  stat_smooth(method = "lm")

melt_glm <- lm(SWE_melt ~ SWE + Mean_Temp + watershed, data = temp_melt_merge)
  summary(melt_glm)
  
# Looks like no significant differences between watersheds, so stick with simpler GLM
  
melt_glm2 <- lm(SWE_melt ~ SWE + Mean_Temp, data = temp_melt_merge)
  summary(melt_glm2)
  
```

Now see how well model predicts daily snow melt in next year  

```{r temp_melt_pred}
daily_temp2016 <- load_csv_from_googledrive("10_0uB8Qxv-AOZbI0MjjHgSnmeP7397D2")
daily_snow_melt2016 <- load_csv_from_googledrive("10ATgwIxJ8cL_A9JcG9N3YajH8E4E087y")

temp_melt_merge2016 <- daily_temp2016 %>% 
  left_join(daily_snow_melt2016, by = c("Date" = "Date",
                                        "watershed" = "watershed"))

temp_melt_merge2016 <- temp_melt_merge2016 %>% 
  mutate(swe_melt_pred = predict.lm(melt_glm2, newdata = temp_melt_merge2016))

temp_melt_merge2016 %>% 
  ggplot(aes(x = swe_melt_pred, y = SWE_melt)) +
    theme_classic() + 
    geom_point() +
    stat_smooth(method = "lm")

temp_melt_merge2016 %>% 
  
  ggplot(aes(x = Date, y = SWE_melt)) + 
    theme_classic() +
    geom_line() +
    geom_line(aes(x = Date, y = swe_melt_pred), lty = 2) +
    facet_wrap(~watershed)
```

